Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
Move a figure to save some space (thanks toon). Actually reference another figure (thanks armin).
Benchmark Psyco

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. To apply hints one needs a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.4 &4.96 $\pm$ 0.05 &4.73\\
\hline
\end{tabular}

@@

To test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. While the function may seem a bit arbitrary, executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations, like indexing into the tuple, addition
and comparison of \texttt{i}. We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.4 (the reference implementation of
Python).

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. To achieve this, the tracer
traces through the whole Python dispatching machinery, automatically inlining
the relevant fast paths.

>>>>

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
parts of the interpreter with hints applied. To apply hints one needs a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

To test the technique on a more realistic example, we did some
preliminary benchmarks with PyPy's Python interpreter. The function we
benchmarked as well as the results can be seen in Figure
\ref{fig:bench-python}. While the function may seem a bit arbitrary, executing it is
still non-trivial, as a normal Python interpreter needs to dynamically dispatch
nearly all of the involved operations, like indexing into the tuple, addition
and comparison of \texttt{i}. We benchmarked PyPy's Python interpreter with the
JIT disabled, with the JIT enabled and
CPython\footnote{\texttt{http://python.org}} 2.5.2 (the reference implementation of
Python). In addition we benchmarked CPython using Psyco 1.6
\cite{rigo_representation-based_2004}, a specializing JIT compiler for Python.

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. To achieve this, the tracer
traces through the whole Python dispatching machinery, automatically inlining
the relevant fast paths. However, the manually tuned Psyco still performs a lot
better than our prototype.
